---
title: "SEtools"
author:
- name: Pierre-Luc Germain
  affiliation: D-HEST Institute for Neurosciences, ETH Zürich & Laboratory of Statistical Bioinformatics, University Zürich
package: SEtools
output:
  BiocStyle::html_document
abstract: |
  Description of your vignette
vignette: |
  %\VignetteIndexEntry{SEtools}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Getting started

The `r Rpackage("SEtools")` package is a set of convenience functions for the _Bioconductor_ class `r Biocpkg("SummarizedExperiment")`. It facilitates merging, melting, and plotting `SummarizedExperiment` objects. To showcase the main functions, we will use an example object which contains (a subset of) whole-hippocampus RNAseq  of mice after different stressors:

```{r}
library(SummarizedExperiment)
library(SEtools)
data("SE")
SE
```

## Heatmaps

The `sehm` function simplifies the generation of heatmaps from `SummarizedExperiment`. It uses `r CRANpkg("pheatmap")`, so any argument supported by it can in principle be passed:

```{r}
g <- c("Egr1", "Nr4a1", "Fos", "Egr2", "Sgk1", "Apold1", "Arc", "Dusp1", "Fosb", "Sik1")
sehm(SE, genes=g)
sehm(SE, genes=g, scale="row")
```

Annotation from the object's `rowData` and `colData` can be plotted simply by specifying the column name (some will be shown by default if found):
```{r}
sehm(SE, genes=g, scale="row", anno_rows="meanTPM")
```

These can also be used to create gaps:
```{r}
sehm(SE, genes=g, scale="row", anno_rows="meanTPM", gaps_at="Condition")
```

Heatmaps from multiple SE can be created either by merging the objects, or using the `crossHm` function, which uses the `r CRANpkg("ComplexHeatmap")` pacakge:

```{r}
crossHm(list(se1=SE, se2=SE), g, acolors = list(Condition=c(Homecage="green", "Handling"="orange", "Restraint"="red", "Swim"="blue")))
```


## Merging SEs

```{r}
se1 <- SE[,1:10]
se2 <- SE[,11:20]
se3 <- mergeSEs( list(se1=se1, se2=se2) )
sehm(se3, genes=g)
```

By default, row z-scores are calculated for each object when merging. This can be prevented with:
```{r}
se3 <- mergeSEs( list(se1=se1, se2=se2), what="asis")
sehm(se3, genes=g)
```

## Melting SE

```{r, fig.cap="An example ggplot created from a melted SE."}
d <- meltSE(SE, genes=g[1:4])
head(d)
suppressPackageStartupMessages(library(ggplot2))
ggplot(d, aes(Condition, counts)) + geom_violin() + facet_wrap(~feature, scale="free")
```


# Session info {.unnumbered}

```{r sessionInfo, echo=FALSE}
sessionInfo()
```
